---
title: imdtools 📦
format:
  gfm:
    preview-mode: raw
---


Tools for Indices of Multiple Deprivation (IMD) data in R.

## Installation and usage

To install:

```r
pak::pak("The-Strategy-Unit/imdtools")
```

```{r}
library(imdtools)
```

To generate a lookup table of IMD2025 ranks and deciles for all English LSOAs:

```r
get_imd_lookup()
```

This table is created from "File 1" on the [GOV.UK IMD2025 webpage][govimd].

[govimd]: https://www.gov.uk/government/statistics/english-indices-of-deprivation-2025

To generate a table of the LSOA-level transformed scores for each of the 7
IMD domains:

```r
get_transformed_scores()
```

This table is created from "File 9" on the [GOV.UK IMD2025 webpage][govimd].


To re-create the published IMD scores by weighting and combining the
transformed domain scores, use the `calculate_imd_score()` function:

```r
get_transformed_scores() |>
  calculate_imd_score(domains = choose_domains())
```

`choose_domains()` helps you select a subset of the seven domains should you
wish to do so, in order to calculate an amended version of the IMD.
This follows the guidance in the [IMD Research Report (PDF)][resrep], Appendix
B:

>  It is possible to use the component domains to produce alternative measures
> of deprivation at LSOA, based on different domain weights than are used in the
> IMD.
>
> For example, health researchers may want to use the IMD as a factor to help
> explain the variation in health outcomes across a sample of areas or
> individuals.
> To exclude the effect of the Health Deprivation and Disability Domain, they
> may want to use a modified measure of deprivation in their statistical
> analysis, with the Health Deprivation and Disability Domain weight set to
> zero.


[resrep]: https://www.gov.uk/government/publications/english-indices-of-deprivation-2025-research-report

This exclusion can be achieved by doing:

```r
get_transformed_scores() |>
  calculate_imd_score(domains = choose_domains(include_health = FALSE))
```

This is equivalent to (but more straightforward than) doing:

```r
new_weights <- c(rep(0.225, 2), 0.135, 0, rep(0.093, 3)) # element 4 set to 0
get_transformed_scores() |>
  calculate_imd_score(weights = new_weights)
```
_(NB the vector must stay length 7 - replace a value by 0; don't _omit_ it)._


Further functions may be added in future if they are thought to be helpful.

## Discrepancy between calculated and published scores and ranks

If we do:

```{r}
calculated_data <- get_transformed_scores() |>
  calculate_imd_score()

overall_scores <- get_overall_scores()
common_cols <- c("lsoa21cd", "lsoa21nm", "lad24cd", "lad24nm")
imd_lookup <- get_imd_lookup() |>
  dplyr::left_join(overall_scores, common_cols) |>
  dplyr::arrange(.data[["imd_rank"]])
```


```{r}
#| echo: false
mismatch <- sum(imd_lookup[["lsoa21cd"]] != calculated_data[["lsoa21cd"]])
```

```{r}
sum(imd_lookup[["lsoa21cd"]] != calculated_data[["lsoa21cd"]])
```


```{r}
#| echo: false
dec_mism <- sum(imd_lookup[["imd_decile"]] != calculated_data[["imd_decile"]])
```

```{r}
sum(imd_lookup[["imd_decile"]] != calculated_data[["imd_decile"]])
```

we can see that `r mismatch` LSOAs are in a different row/rank to what is
expected from the published ranks.

And in `r dec_mism` cases this even causes the LSOA to be allocated to a
different decile.

### Published instructions for calculating an overall score from transformed domain scores and weights

This approach ought to be the correct one, following the published guidance in
the IMD Research Report, which makes clear:

> The standardised domain scores have been standardised by ranking and then
> transformed to an exponential distribution.
> *These standardised domain scores have been published to be used as the basis
> for users to combine the domains together using different weights.*</br />
> *[emphasis added]*<br />
> [Reconstituting the IMD overall score] can be achieved... using the following
> equation:
>
> Income Deprivation Domain x domain-weight +<br />
> Employment Deprivation Domain x domain-weight + ... etc.

Simply: each transformed (standardised) domain score is multiplied by its
appropriate domain weight, then all the results are summed together.

This is the approach implemented in our function `calculate_imd_score()`.

### Explaining the discrepancy - and correcting it?

One thought is to replace the published value for the weightings for domains 5,
6 and 7 with a slightly amended value: `(0.28 / 3)` instead of `0.093` exactly.

This would make the seven weights add up to 100% exactly, and it seems not an
unreasonable assumption that for this reason the fractional value might have
been the one actually used by the creators of the IMD.

However, amending these three weights makes the problem worse instead of better:

```{r}
# use 0.28/3 instead of 0.093
new_weights <- c(rep(0.225, 2), rep(0.135, 2), rep(0.28 / 3, 3))
new_calculated_data <- get_transformed_scores() |>
  calculate_imd_score(weights = new_weights)
```

```{r}
sum(imd_lookup[["lsoa21cd"]] != new_calculated_data[["lsoa21cd"]])
```


The lesson here is that the documentation is actually accurate!

The IMD scores are published to an accuracy of 3 decimal places.
The values we get from `calculate_imd_score()` are not rounded.
In the majority of cases, if we round our calculated scores to 3dp, they are
within 0.001 of the published values.

```{r}
calc2 <- calculated_data |>
  dplyr::rename_with(\(x) sub("^imd", "calc_imd", x)) |>
  # round to 3dp
  dplyr::mutate(dplyr::across("calc_imd_score", \(x) round(x, 3)))

# Number of LSOAs where rounded calculated score is equal to published score
imd_lookup |>
  dplyr::left_join(calc2, common_cols) |>
  dplyr::filter(imd_score == calc_imd_score) |>
  nrow()
```

```{r}
# Number of LSOAs where rounded calculated score within 0.001 of published score
imd_lookup |>
  dplyr::left_join(calc2, common_cols) |>
  dplyr::filter(abs(imd_score - calc_imd_score) <= 0.001) |>
  nrow()
```

This is a great reassurance that our calculation process is fundamentally
correct!
There are just some discrepancies.
And it's hard to tell where these arise.

Might there be an issue with how the published IMD scores have been rounded?

The value of `r mismatch` seems suspiciously not too far off 10% of the total
number of LSOAs in England.

If the published values have been rounded to 3dp from more precise calculated
values, we would expect about 10% of those more precise values to have a 4th dp
value of 5.

In the R `round` function, a 5 is "rounded to even", not simply rounded up.
That means that

```{r}
round(0.1235, 3)
```

produces `0.124` in the way that rounding 5s up would produce.
However,

```{r}
round(0.1225, 3)
```

doesn't return `0.123`, but is instead "rounded to even", returning `0.122`.

If the IMD creators used different software that uses rounding up, could this
explain our discrepancies?

Unfortunately not; not based on our calculated values, anyway.
There is no pattern that suggests rounding up instead of rounding to even would
explain the discrepancies.

What about the ranking of LSOAs with tied scores?
If we have 2 (or more) LSOAs with the same overall score (when expressed to
3dp), perhaps residual ordering from an original table of LSOAs might be
retained in the production of overall ranks.
Whereas ranking based on our more precise (more than 3dp) calculated values
might break the apparent ties in a different way.

Again, however, on inspection there seems to be no reason to think that an
initial ordering of LSOAs can explain the discrepancies.

It might be possible to hack our calculated values so that they more closely
instantiate the published ranks

Although we believe our calculation function is correctly implementing the
published method, we may be able to

Through some semi-automated trial and error, we find that a bi-directional
adjustment value of `0.000167` can be applied programmatically.
This correction reduces the number of LSOAs with a discrepancy in rank compared
to their official published ranks.

```{r}
fix_calculated_data <- function(calculated_data, imd_lookup, adj = 0.000167) {
  common_cols <- c("lsoa21cd", "lsoa21nm", "lad24cd", "lad24nm")
  calculated_data |>
    dplyr::rename_with(\(x) sub("^imd", "calc_imd", x)) |>
    dplyr::left_join(imd_lookup, common_cols) |>
    dplyr::mutate(dplyr::across("calc_imd_score", \(x) {
      dplyr::case_when(
        # if our calculated rank is too low, that means our score was too high
        .data[["calc_imd_rank"]] < .data[["imd_rank"]] ~ x - adj,
        # if our calculated rank is too high, that means our score was too low
        .data[["calc_imd_rank"]] > .data[["imd_rank"]] ~ x + adj,
        .default = x
      )
    })) |>
    dplyr::select(!tidyselect::starts_with("imd")) |>
    dplyr::arrange(dplyr::desc(.data[["calc_imd_score"]])) |>
    dplyr::mutate(
      calc_imd_rank = dplyr::row_number(),
      calc_imd_decile = as.factor(dplyr::ntile(n = 10))
    ) |>
    dplyr::rename_with(\(x) sub("^calc_imd", "imd", x))
}

fixed_calculated_data <- fix_calculated_data(calculated_data, imd_lookup)

```

```{r}
#| echo: false
mismatch <- sum(imd_lookup[["lsoa21cd"]] != fixed_calculated_data[["lsoa21cd"]])
```

```{r}
sum(imd_lookup[["lsoa21cd"]] != fixed_calculated_data[["lsoa21cd"]])
```


After applying our illicit adjustment hack, we can now see that just
`r mismatch` LSOAs have a different rank to what is expected.


```{r}
rounded_data <- fixed_calculated_data |>
  dplyr::rename_with(\(x) sub("^imd", "calc_imd", x)) |>
  dplyr::mutate(dplyr::across("calc_imd_score", \(x) round(x, 3))) |>
  dplyr::left_join(imd_lookup, common_cols) |>
  dplyr::filter(imd_rank != calc_imd_rank) |>
  dplyr::filter(imd_score != calc_imd_score)


# rounded_data |>
#   dplyr::mutate(dplyr::across(tidyselect::ends_with("imd_score"), as.character)) |>
#   View()
```

```{r}
#| echo: false
n_diffs_rounded <- nrow(rounded_data)
```

And of those still with incorrect ranks, there are now `r n_diffs_rounded`
rows with differences between the calculated scores (when rounded to 3dp) and
the published scores.


### Conclusions

1. Scores calculated by `calculate_imd_score()` are in general close enough
  (within 0.001) of the published overall score to validate that the
  calculation process in the function is correctly implemented.
2. If the aim is to reproduce the published overall LSOA scores and ranks, the
  functions here do not succeed in completely accurately generating the expected
  scores.
  A correcting function, `fix_calculated_data()`, which has no
  theoretical basis, may be applied to the calculated data.
  This function is currently only found in the README file, not within the
  package namespace.
  This function reduces the error rate in calculated scores from approximately
  10% to approximately 1%.
  Most importantly, the correcting function reduces the magnitude of the
  discrepancies so that in this context the number of LSOAs assigned to the
  wrong decile was reduced from 15 to zero.
  However this function relies on the published table of ranks as the basis for
  comparison in the application of the adjustment quantity, and is therefore
  only useful when trying to replicate the overall scores and ranks.
  In which case one might as well just use the original data!
  If calculating scores using a custom domain selection or custom weights,
  `fix_calculated_data()` is not applicable, because we have no authoritative
  pre-calculated table to compare calculated results against.
  In these cases, the user should just trust that `calculate_imd_score()` will
  provide "as good as can be expected" results in what is after all a process
  (production of the IMD) that inevitably contains a number of assumptions and
  approximations.
3. It remains unclear why the ~10% discrepancy rate between published and
  calculated overall scores exists.
  It seems clear that the default weightings stored in `imd_weights()` are
  accurate.
  If IMD overall scores (and/or transformed scores) were available with a
  precision greater than 3 decimal places, we might be able to investigate in a
  more detailed way where the discrepancies are being introduced, and how.


## Reporting problems

Please use GitHub Issues to report any problems or make suggestions for further
development.
